---
title: "preprocess_data"
author: "Deiver Suárez Gómez - Alibeth Luna"
date: "2/20/2020"
output:
  word_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##1. Establishment of the working directory

```{r}

setwd("/Users/deiversuarezgomez/Desktop/entrenamiento")  
```

##1. Install and load GEOquery package onto R.

```{r}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("GEOquery")



#Load the GEOquery
library(GEOquery)
library(tidyr)
library(stringr)
library(dplyr)
```

##2. Download a GSE dataset. In particular, an endometriosis dataset intending to study human endometriosis vs normal endometrium using transcriptional profiling.
.
```{r}
gset <- getGEO("GSE120584", GSEMatrix =TRUE, getGPL=FALSE) 
```

##3. Convert from Large List format to Large ExpressionSet. Basically, we will be selecting the matrix with information which is in the first item of the Large List.

```{r}
summary(gset)

gse=gset[[1]]
```

##4. Extract the microarray expression values.
```{r}
exp=exprs(gse)
dim(exp)
head(exp)

row.names(exp)
```

##5. Extract phenotypic information from these samples.
```{r}
pheno = pData(gse)
head(pheno)

phdata = pheno[,c("geo_accession","diagnosis:ch1","Sex:ch1")]


table(phdata$`diagnosis:ch1`)

dummy = phdata[phdata$`diagnosis:ch1` %in% c("AD", "NC"),] #aqui se filtra por ALN y control

table(dummy$`Sex:ch1`)

dummy$"diagnosis:ch1" = gsub("AD","disease", dummy$"diagnosis:ch1")
dummy$"diagnosis:ch1" = gsub("NC","control", dummy$"diagnosis:ch1")

dummy$"Sex:ch1" = gsub("Female","female", dummy$"Sex:ch1")
dummy$"Sex:ch1" = gsub("Male","male", dummy$"Sex:ch1")

phdata = dummy




names(phdata)[2]="disease.state"
names(phdata)[3]="Sex"
```

```{r}
result <- phdata %>%
  group_by(disease.state, Sex) %>%
  summarise(count = n()) %>%
  ungroup()

print(result)
```

##5. Download the GPL file and convert into table format.
```{r,warning=FALSE}

gpl = getGEO("GPL21263")
gpltable = Table(gpl)   #extraer Table(funcion particula GEOquerry para extraer la tabla gpl)

#all(gpltable$ID == row.names(exp))

dim(gpltable)         
gpltable
```

##6. To convert from probes to genes, let's start by merging probe's annotation information.
```{r}
newGPL = gpltable[,c("ID","miRNA")]

mdata = merge(newGPL,exp,by.x="ID",by.y=0)

dummy2 = mdata[,-1]

dummy = separate(data = dummy2, col="miRNA",sep=",", into = c('x1','x2'),remove = TRUE)

dummy = dummy[,-2]

names(dummy)[1]="miRNA"

mdata = dummy
```


##7. Let's verify how many probes were mapped for each identified gene symbol.
```{r}
p2g = data.frame(table(mdata$`miRNA`)) #cuanto de esos genesimbol se repiten
table(p2g$Freq)
```

##8.Choose the probe with maximum expression as the main probe to characterize the gene expression of a gene symbol. To perform this operation we need the commmand aggregate from tidyr.
```{r}
mdata$"miRNA" = gsub("-","_",mdata$"miRNA")

gdata = mdata #elimina la primera fila es las filas que estan en blanco 
dim(gdata)
gdata[1:50,1:10]
```

##9. Another data preprocessing issue, special caracters between gene aliases. In this step, we will remove those gene aliases from the expression table.
```{r, warning=FALSE}
library(stringr)
gexp = gdata
```


##11.Merge the gene expression file (gexp) with the patients information (psamples)
```{r}

dummy = data.frame(t(gexp[,-1])) #qito el nombre de los genes para poder hacer la transpuesta


names(dummy) = gexp[,1]  #nombre d elos genes

final = merge(phdata,dummy,by = 0)
final = final[,-1]

```


```{r}
ALZ3_sex_final_GSE195872 = final[,c(1,3)]

write.csv(ALZ3_sex_final_GSE195872, file = "AD_sex_GSE120584.csv", row.names = FALSE)
```



```{r}
ALZ3_GSE195873 = final[,-3]

write.csv(ALZ3_GSE195873, file = "AD_GSE120584.csv", row.names = FALSE)
```


